### 2023-07-11 abstract.tex

\begin{abstract}
Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions entirely meaningless. 
This approach is insufficient as the basis for language servers for modern programming environments, which are expected to sensibly localize any type errors and recover, i.e. continue to provide a variety of downstream semantic services. 
This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery: the marked lambda calculus. In particular, we define a gradual type system for expressions with marked errors and holes, together with a total procedure for marking unmarked expressions. We mechanize the metatheory of the marked lambda calculus in Agda and implement it, scaled up, as the new basis for Hazel, a full-scale live functional programming environment. 
As a secondary contribution, we show how total marking resolves 
the problem of undefined behavior in the Hazelnut structure editor calculus
(though marking itself does not require a structure editor).

The marked lambda calculus supports local type inference, i.e. it is bidirectionally typed. Constraint-based type inference notoriously complicates error localization and recovery. We approach this problem by deploying constraint solving as a type-hole-filling layer atop our gradual bidirectionally typed core. The unification algorithm we contribute localizes errors arising from inconsistent constraints to type holes with traced  provenances, rather than attempting to localize them in an \emph{ad hoc} manner to particular expressions. The user can choose how to shift these errors to expressions by selecting from suggested partial solutions, or allow the unresolved type holes to recover gradually. We implement this type hole inference system in Hazel. 
\end{abstract}

>>>>

\begin{abstract}
Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions formally meaningless. 
This approach is insufficient as the basis for language servers driving modern programming environments, which are expected to recover from simultaneously localized errors and continue to provide a variety of downstream semantic services. 
This paper addresses this problem, contributing the first comprehensive formal account of total type error localization and recovery: the marked lambda calculus. In particular, we define a gradual type system for expressions with marked errors, which operate as non-empty holes, together with a total procedure for marking arbitrary unmarked expressions. We mechanize the metatheory of the marked lambda calculus in Agda and implement it, scaled up, as the new basis for Hazel, a full-scale live functional programming environment with, uniquely, no meaningless editor states. 
% As a secondary contribution, we show how total marking resolves 
% the problem of undefined behavior in the Hazelnut structure editor calculus
% (though marking itself does not require a structure editor).

The marked lambda calculus is bidirectionally typed, so localization decisions are systematically predictable based on a local flow of typing information. 
Constraint-based type inference can bring more distant information to bear in discovering inconsistencies but this notoriously complicates error localization. We approach this problem by deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed core. Errors arising from inconsistent unification constraints are localized exclusively to type and expression holes, i.e. the system identifies unfillable holes using a sytem of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions. The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system. We implement this type hole inference system in Hazel. 
\end{abstract}

### 2023-07-11 intro.tex

These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook,pfpl},
assigns meaning only to well-typed programs. 
If a type error appears \emph{anywhere}, the program is formally meaningless \emph{everywhere}.

This gap problem has prompted considerable practical interest in 
(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
(2) \textbf{type error recovery}: mechanisms that allow the language server to optimistically recover from a localized type error 
and continue on to locate other errors and provide downstream semantic services, 
ideally at every location in the program and with minimal degradation in service.
Essentially all widely-used programming systems have some support for type error localization, 
reporting error locations in compiler error messages 
or directly in the editor via markings decorating the localized errors. Developers attend primarily to these error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
Many systems also attempt recovery in certain situations, discussed below.

However, type error localization and recovery mechanisms have developed idiosyncratically, 
in part as folklore amongst language and tool implementors. 
Different type checkers or language servers \cite{barros2022editing,bour2018merlin}, even for the same language, localize and recover from type errors in different ways, 
with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.

Consider, the example program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
\vspace{-3px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot.png}
\end{center}
\vspace{-3px}
A type checker with no support for error localization or recovery---%
as students in an undergraduate course might write---%
would simply report that this program is ill-typed. 
A more practical approach, and one common even in production type checkers, 
is to localize the first error that causes local type inference to fail and emit an explanatory error message.
In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then stop.

An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018merlin}, would 
 be tasked with continuing past this first error.
 The general difficulty here is that there is now missing semantic information, namely the type of \li{f}, that 
 the bidirectional type system as specified 
 would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against.
 Intuitively, however, it seems that this type is actually \emph{unnecessary} to make an error localization decision about \li{y}: 
it is unbound, so a second error can be simultaneously localized despite the missing information. Elm stops here.

To recover further, we might choose to also ignore the bidirectional type system's demand that the guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Int} and \li{Bool}. 
There are several ways to localize this error. 
One approach, taken e.g. by \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g. 
by observing recent editing history \cite{steady-typing}, 
or training a machine learning model \cite{SeidelBlame}. 
The other approach, which Hazel takes, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated above. When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{potter2020hazel}---shows:

\begin{center}
  \includegraphics[width=0.9\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
\end{center}

This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
If localization had assumed that the \li{then} branch correct, as for example in the Elm type checker, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
If the \li{else} branch was chosen, then \li{x : Int} and an error would not be reported. 
If the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error here, 
so we choose not to report an error in Hazel.
No matter how localization handles the left operand, we would like to be able to recover and 
localize the type inconsistency on the right operand (the \li{+} operator is integer addition in Hazel).

This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to whether they make \emph{ad hoc} guesses about intent;  
(2) when combined with error recovery, localization decisions can influence downstream localization decisions; and 
(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of the topic.

>>>>

These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook},
assigns meaning only to well-typed programs. 
If a type error appears \emph{anywhere}, the program is formally meaningless \emph{everywhere}.

This gap problem has prompted considerable practical interest in 
(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
(2) \textbf{type error recovery}: mechanisms that allow the language server to optimistically recover from a localized type error 
and continue on to locate other errors and provide downstream semantic services, 
ideally at every location in the program and with minimal degradation in service.
Essentially all widely-used programming systems have some support for type error localization, 
reporting error locations in compiler error messages 
or directly in the editor via markings decorating the localized errors. Developers are known to attend to these error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
Many systems also attempt recovery in certain situations, discussed below. 
However, type error localization and recovery mechanisms have developed idiosyncratically, 
in part as folklore amongst language and tool implementors. 
Different type checkers or language servers \cite{barros2022editing,bour2018merlin}, even for the same language, localize and recover from type errors in different ways, 
with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.

Consider, the example ill-typed program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
\vspace{-3px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot.png}
\end{center}
\vspace{-3px}
A type checker with no support for error localization or recovery---%
as students in an undergraduate course might write---%
would simply report that this program is ill-typed. 
A more practical approach, and one common even in production type checkers, 
is to localize the first error that causes local type inference to fail and emit an explanatory error message before terminating.
In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then terminate.

An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018merlin}, would 
 be tasked with continuing past this first error.
 The general difficulty here is that there is now missing semantic information, namely the type of \li{f}, that 
 the bidirectional type system as specified 
 would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against.
 Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
it is unbound, so a second error can be simultaneously localized despite the missing information about \li{f}. Elm's compiler stops here.

To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Int} and \li{Bool}. 
There are several ways to localize this error. 
One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g. 
by observing recent editing history \cite{steady-typing}, 
or training a machine learning model \cite{SeidelBlame}. 
The other approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ, as indicated above. When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{potter2020hazel}---shows:

\begin{center}
  \includegraphics[width=\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
\end{center}

This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
If localization had assumed that the \li{then} branch correct, as for example in Elm and \li{merlin}, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
If the \li{else} branch was chosen, then \li{x : Int} and an error would not be reported. 
If the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error here, 
so we choose not to report an error in Hazel.
No matter how localization handles the left operand, we would like to be able to recover and 
localize the type inconsistency on the right operand (the \li{+} operator is integer addition in Hazel).

This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to whether they make \emph{ad hoc} guesses about intent;  
(2) when combined with error recovery, one localization decision can influence downstream localization decisions; and 
(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of the topic.
